Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages

### 2014-03-04 docs/acl-2014/paper.tex
Merged previous versions.
Added comments about copyright.

At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com/}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century. However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

@@

The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags. Each XML file is saved under a filename that matches the respective language code. While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation. [Huh? Should this really happen? - the codes are given on their website and look valid to me...]

\begin{minipage}{\columnwidth}

a. \quad o lesu mai \\
\indent \qquad 2sg return here \\

\indent \qquad `\emph{You return here.}' \\
\centerline{Figure 1: IGT from ODIN.}
\end{minipage}

\begin{minipage}{\columnwidth} %Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igit>
\end{lstlisting} 
\centerline{Figure 2: Fijian IGT from ODIN.}
\end{minipage}
\\ \\
\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss. From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2. Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\centerline{Figure 3: Yimas IGT from ODIN.}
\end{minipage}
\\ \\
\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs. In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies. In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.

@@

The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages. We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.

[The following discussion is far too technical.]

@@

\noindent The \emph{`Useful foreign phrases'} page contains parallel phrases in embedded within the \texttt{<th><tr>...</tr><th>} tags. The English phrase and foreign language phrase are encoded separately with \texttt{<td>...</td>} tags. Figure 4 shows an instance of an Omniglot \emph{Useful foreign phrase} page. Since the HTML page was designed as WYSIWYG, several pages have non-textual elements within the \texttt{<tr>...</tr>} tags for aesthetics. For example, the non-breaking space, \texttt{<td>\&nbsp;</td>} in Fig. 4; the extracted text was properly converted into plaintext format using the HTMLParser\footnote{http://docs.python.org/2/library/htmlparser.html} module, during which the non textual elements would have been cleaned.

The same process of crawling URLs and cleaning was performed on the \emph{`Tower of Babel'} pages and the only difference was that the texts were embedded in a \texttt{<ol><li>...</li></ol>} tags instead.

\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<th>
  ...
  <tr>
    <td>I don't understand</td>
    <td>No appo cumpresu nutta</td>
    <td>&nbsp;</td>
  </tr>
  <tr>
    <td>I don't understand</td>
    <td>Non d'isco</td>
    <td>&nbsp;</td>
  </tr>
  ...
<th>
\end{lstlisting} 
\centerline{Figure 4: A \emph{Useful Foreign Phrase} page}
\end{minipage}
\smallskip

@@

The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages. Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format. Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues. The first four lines of each file records metadata, and the rest is the translation of the UDHR. This dataset was extremely clean, and simply required segmentation into sentences.

@@

Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation needed}}). [It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we used the wp-download\footnote{https://github.com/babilen/wp-download} tool, adapted to an expanded set of languages.

%To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical][Such script has not been implemented]

%\smallskip
%\noindent \texttt{\$ wget http://download.wikimedia.
%org/\$Iwiki/latest/itwiki-latest-
%pages-articles.xml.bz2}
%\medskip

\noindent One major issue with using the Wikipedia dumps is the sheer size and extracting text from a glut of Wikipedia markup. To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool. After the conversion into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.

Wikipedia uses its own set of language codes, most of which are in ISO 639-1 or ISO 639-2/3. We automatically converted all of them into ISO~639-3.

\subsection{Leipzig Corpora Collection}

Do we want to mention this?

@@

Creative Commons Licence, etc.

We have made all our resources publically available.\footnote{To maintain anonymity, we have removed details, but, if accepted for publication, instructions will be included in the print version of this paper.}

>>>>

At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century. However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.

@@

The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags. Each XML file is saved under a filename that matches the respective language code. While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation. [Huh? Should this really happen? - the codes are given on their website and look valid to me...]
\\ \\
\begin{minipage}{\columnwidth}

\quad 21 a.\quad o lesu mai \\
\indent \qquad\qquad 2sg return here \\
\indent \qquad\qquad `\emph{You return here.}' \\
\centerline{Figure 1: Fijian IGT from ODIN.}
\end{minipage}

%%
% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
% Also, some of the lines go outside of the column!!
%%

\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>21 a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igt>
\end{lstlisting} 
\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
\end{minipage}
\\ \\
\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss. From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2. Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    Yimas (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
\end{minipage}
\\ \\
\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs. In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies. In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.

@@

The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages. We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages. The urls for these resources are:

@@

@@

The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages. Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format. Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues. The first four lines of each file records metadata, and the rest is the translation of the UDHR. This dataset was extremely clean, and simply required segmentation into sentences.

@@

At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages. Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}. One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup. To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool. After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.

Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3. We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}

@@

Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgarriff, 2002)[@Li, please add this reference to the bibliography file]. Often, free access to language data online does not imply the right to download, retain, remix and redistribute.

ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents. As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site. However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.

The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page\footnote{http://www.omniglot.com/faqs.htm} but the rights of the Tower of Babel snippets from different bible sources remains iffy. For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}. [Ga: actually, about copying information, Omniglot says: \emph{'If you would like to copy or re-use any of the language-related articles, please ask the authors for permission. Contact details of the authors can be found at the bottom of the articles in most cases, if they're not there, please contact me, and I'll try to find them for you.'} - I suspect Adger has been given permission for each translation, and I think we would have to ask every single author if we want to re-distribute the data.  For the tower of babel translations, this could end up a nightmare...]

Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, [Ga: Do we have a reference for this? Particularly distribution] there is no explicit ownership of the UDHR textfiles from the Unicode website. However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part, in any form or by any means, electronic or mechanical, including photocopying, recording or the use of any information storage and retrieval system, except as provided for in the Terms and Conditions of Use of United Nations Web Sites, without permission in writing from the publisher.}'\footnote{http://www.un.org/en/aboutun/copyright/}. [Ga: Actually, it's not quite so bad - if we look at the terms and conditions, it says: \emph{'The United Nations grants permission to Users to visit the Site and to download and copy the information, documents and materials (collectively, “Materials”) from the Site for the User’s personal, non-commercial use, without any right to resell or redistribute them or to compile or create derivative works therefrom, subject to the terms and conditions outlined below, and also subject to more specific restrictions that may apply to specific Material within this Site.'}\footnote{http://www.un.org/en/aboutun/terms/} Specifically for the UDHR, it says: \emph{'If UDHR translations or materials are reproduced, users should make reference to this website as a source by providing a link.'}\footnote{http://www.ohchr.org/EN/UDHR/Pages/Introduction.aspx}  So, it looks like we're okay to use the UDHR as we currently are, but if we want to redistribute it, we need to get their permission in writing.]

The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation. Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.  We have made this resource publically available.\footnote{To maintain anonymity, we do not give details of how to access the data, but, if accepted for publication, instructions will be included in the print version of this paper.}

%%
%To avoid any copyrights infringement, ... After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
%%
